Scoring Patients

### 2016-12-01 Patient Report.Rmd
Mapping genes from the cosmic gene consensus onto the variable regions.

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "15 November 2016"
output: pdf_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)

@@

my_custom_cor <- function(data, mapping, color = I("grey10"), sizeRange = c(1, 7), ...) {
  # get the x and y data to use the other code
  x <- eval(mapping$x, data)
  y <- eval(mapping$y, data)

  ct <- cor.test(x,y)
  sig <- symnum(
    ct$p.value, corr = FALSE, na = FALSE,
    cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("***", "**", "*", ".", " ")
  )

  r <- unname(abs(ct$estimate))
  rt <- format(r, digits=2)[1]

  # since we can't print it to get the strsize, just use the max size range
  cex <- max(sizeRange)

  # helper function to calculate a useable size
  percent_of_range <- function(percent, range) {
    percent * diff(range) + min(range, na.rm = TRUE)
  }

  sig_star_color = color
  if (ct$p.value <= 0.05)
    sig_star_color = I('firebrick')

  # plot the cor value
  p = ggally_text(
    label = as.character(rt), 
    mapping = aes(),
    xP = 0.5, yP = 0.5, 
    size = I(percent_of_range(cex * abs(r), sizeRange)),
    color = color,
    ... ) +

    # add the sig stars
    geom_text(
      aes_string( x = 0.8, y = 0.8 ),
      label = sig, 
      size = I( (percent_of_range(cex * abs(r), sizeRange)/2) ),
      #size = I(cex),
      color = sig_star_color,
      ... ) + 
    # remove all the background stuff and wrap it with a dashed line
    theme_classic() + 
    theme(
      panel.background = element_rect(
        color = color, 
        size = 2,
        linetype = "longdash"
      ), 
      axis.line = element_blank(), 
      axis.ticks = element_blank(), 
      axis.text.y = element_blank(), 
      axis.text.x = element_blank()
    )
}

panel.cor <- function(x, y, digits=2, prefix="", cex.cor) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use= "pairwise.complete.obs"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex <- 0.8/strwidth(txt)

  test <- cor.test(x,y, use= "pairwise.complete.obs")
  # borrowed from printCoefmat
  Signif <- symnum(test$p.value, corr = FALSE, na = FALSE,
                   cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                   symbols = c("***", "**", "*", ".", " "))

  text(0.5, 0.5, txt, cex = cex * r)
  text(.8, .8, Signif, cex=cex, col=2)
}

my_bars<-function(data, mapping, sizeRange=c(1,7),...) {
  # get the x and y data to use the other code
  x <- eval(mapping$x, data)
  y <- eval(mapping$y, data)

  name = as.character(mapping$x)
  ggplot(data=data,mapping=mapping) + geom_histogram(bins=10,col='green4',fill='palegreen',alpha=0.7) + 
    annotate("text",label=name, x=Inf, y=Inf, vjust=1,hjust=1,col=I('grey14'),fontface=2,size=max(sizeRange) )
}

panel.hist <- function(x, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}

@@

## Patient info file
patient.file = grep('patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

@@

patient.data = ddply(patient.info, .(Patient, Status), summarise, years=paste(range(Endoscopy.Year), collapse='-'), 
      n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T))

`r pander(arrange(patient.data, Status), justify='left', caption='Patient information')`

## Gamma = 250
```{r cors, echo=F, message=T, warning=F, fig.height=6, fig.width=6, fig.align='left'}
min.probes=67
gamma2 = 250

cor.scores = as.data.frame(matrix(nrow=length(patient.data$Patient), ncol=3, dimnames=list(patient.data$Patient, c('similar','different','n.probes'))))
for (patient.name in unique(patient.info$Patient)  )  {
  message(patient.name)
  patient.plot.dir = paste(plot.dir, patient.name, sep='/')

@@

  if(nrow(segvals)>0) {

      HC = hclust(dist(t(segvals[,-(1:5)])))
      gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
      #print(gg1)

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,-(1:5)]
      for(c in 1:nrow(normalised.segvals)) {
        normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))
      }

@@

      if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology)

        ## TODO Mistake in earlier version of the patient file caused this will be fixed for the next run
        snames = grep('_1072(5|9)$', patient$Samplename)
        if ( length(snames > 0) ) {
          patient$Samplename[snames] =  paste( sub('-','_', patient$Plate.Index[snames]), '10725_10729', sep='_' )
        }

        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]

        rc = rcorr(as.matrix(sample.normalised.segvals), type='pearson')

        sim = 0 
        diff = 0
        for (i in 1:ncol(rc$r)) {
          len = length(which(rc$P[,i] < 0.05))

          sim = sim + signif( sum(rc$r[which(rc$P[,i] < 0.05 & rc$r[,i] > 0),i ]), 2)
          diff = diff + signif(sum(rc$r[which(rc$P[,i] < 0.05 & rc$r[,i] < 0),i ]),2)
        }  
        cor.scores[patient.name,] = c(sim,diff,nrow(sample.normalised.segvals))

        # pairs = ggpairs( sample.normalised.segvals,
        #                  lower=list(continuous=wrap("points", colour='darkblue', shape=20, alpha=0.3, size=3)),
        #                  diag=list(continuous=wrap(my_bars, sizeRange=c(1,5))),
        #                  upper=list(continuous=my_custom_cor),
        #                  axisLabels='show', title=patient.name, columnLabels=paste(paste(patient$Endoscopy.Year, ' (',patient$Pathology, ')', sep=''), patient$Plate.Index, sep='\n') )
        # 
        # print(pairs  + theme( # this doesn't work...
        #   text = element_text(size=20, face='bold')
        # ))

      }
    }
  }
cor.scores$Status = patient.data[which(patient.data$Patient %in% rownames(cor.scores)), 'Status']

```
\pagebreak

```{r scores, echo=F, message=T, warning=F, fig.height=4,fig.width=4,fig.align='left'} 
pander(cor.scores[order(cor.scores$Status, cor.scores$similar),], justify='left')

ggplot(cor.scores, aes(x=similar,y=different, color=Status)) + geom_point()

ggplot(cor.scores, aes(y=abs(similar+different), x=c(1:nrow(cor.scores)), color=Status)) + geom_point()

@@

>>>>

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "15 November 2016"
output: word_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)

library(tibble)

@@

@@

## Patient info file
patient.file = grep('All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

@@

patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year, Pathology)

sum.patient.data = ddply(patient.info, .(Patient, Status), summarise, years=paste(range(Endoscopy.Year), collapse='-'), 
      n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T))

#pander(arrange(patient.data, Status), justify='left', caption='Patient information')
```

## Gamma = 250
```{r cors, echo=F, message=T, warning=F, fig.height=6, fig.width=6, fig.align='left'}
## Thresholds
min.probes=67 # equiv to 1MB
gamma2 = 250
sd.threshold = 0.08
gain.threshold = 1.1
loss.threshold = 0.9

patient.data = list()

for (patient.name in unique(patient.info$Patient)  )  {
  message(patient.name)
  patient.plot.dir = paste(plot.dir, patient.name, sep='/')

@@

  # ORDER segvals by endoscopy year etc
  cur.pt = subset(patient.info, Patient == patient.name)
  segvals[, c(colnames(segvals[1:5]), cur.pt$Samplename)]

  patient.data[[patient.name]][['seg.vals']] = segvals
  patient.data[[patient.name]][['info']] = cur.pt

  if(nrow(segvals)>0) {
      HC = hclust(dist(t(segvals[,-(1:5)])))
      gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
      #print(gg1)

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,-(1:5)]
      for(c in 1:nrow(normalised.segvals)) {
        normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))
      }
      patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

      # Variable Regions
      sds = apply(segvals[,-(1:5)],1,sd)
      means = apply(segvals[,-(1:5)],1,mean)
      CofV = sds/means

      variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
      variable.regions = segvals[variable.region.indices, patient.data[[patient.name]]$info$Samplename] 
      patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )

@@

      if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology)
        }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

## Note the number of probes per patient/sample
```
\pagebreak

```{r scores, echo=F, message=T, warning=F, fig.height=4,fig.width=4,fig.align='left'} 
#pander(cor.scores[order(cor.scores$Status, cor.scores$similar),], justify='left')

#ggplot(cor.scores, aes(x=similar,y=different, color=Status)) + geom_point()

#ggplot(cor.scores, aes(y=abs(similar+different), x=c(1:nrow(cor.scores)), color=Status)) + geom_point()
```

## Variable Regions
```{r var_reg, echo=F, message=F, warning=F}
rm(prog.regions, np.regions)
for (patient.name in sum.patient.data$Patient) {
  patient = patient.data[[patient.name]]

  sum.patient.data[which(sum.patient.data$Patient == patient.name), 'var.regions'] = nrow(patient$variable.regions)

  if (subset(sum.patient.data, Patient == patient.name)$Status == 'P') {
    if (!exists('prog.regions')) {
      prog.regions = patient$variable.regions[,c('chrom','start.pos','end.pos')]
    } else {
      prog.regions = rbind(prog.regions, patient$variable.regions[,c('chrom','start.pos','end.pos')])
    }
  } else {
    if (!exists('np.regions')) {
      np.regions = patient$variable.regions[,c('chrom','start.pos','end.pos')]
    } else {
      np.regions = rbind(np.regions, patient$variable.regions[,c('chrom','start.pos','end.pos')])
    }
  }
}

pander(sum.patient.data %>% group_by(Status) %>% summarise(mean.num.regions = mean(var.regions), sd.num.regions = sd(var.regions)), justify='left')

prog.regions$length = prog.regions$end.pos - prog.regions$start.pos
np.regions$length = np.regions$end.pos - np.regions$start.pos

chrs = read.table('hg19_chrom.txt',sep='\t', header=T)
prog.regions = merge(prog.regions, chrs, by.x='chrom', by.y='chr') 
np.regions = merge(np.regions, chrs, by.x='chrom', by.y='chr')

prog.regions$chr.ratio = prog.regions$length.x/prog.regions$length.y
np.regions$chr.ratio = np.regions$length.x/np.regions$length.y

ttR = t.test(prog.regions$chr.ratio, np.regions$chr.ratio)

np = np.regions %>%
  group_by(chrom) %>%
  summarise(lengths = sum(length.x))

prog = prog.regions %>%
  group_by(chrom) %>%
  summarise(lengths = sum(length.x))

m = merge(prog, np, by='chrom', all.x=T)
tt = t.test(m$lengths.x, m$lengths.y)

plot(hclust( dist(chrs$length)), main='Chromosome cluster by length')

## Grouping the chromosomes by length 1-12, 13-22

t.test(subset(np.regions, chrom %in% c(1:12))$chr.ratio, subset(np.regions, chrom %in% c(13:22))$chr.ratio)

t.test(subset(prog.regions, chrom %in% c(1:12))$chr.ratio, subset(prog.regions, chrom %in% c(13:22))$chr.ratio)

df = data.frame( 'short' = rbind( 'np' = sum(table(np.regions$chrom)[1:12]),
       'p' = sum(table(prog.regions$chrom)[1:12])),
                 'long' = rbind( 'np' = sum(table(np.regions$chrom)[13:22], na.rm=T),
       'p' = sum(table(prog.regions$chrom)[13:22], na.rm=T)))

ftL = fisher.test(df)
```

So it appears that adjusting for chromosome length (looking at the ratio of the chromosome rather than the raw bp) there is still a significant difference between progressers and non: t-test p-value `r ttR$p.value`

There appears to be no enrichment based on the longer vs shorter chromosomes either (fisher `r ftL$p.value`)

```{r cor, echo=F, message=T, warning=F, fig.height=4,fig.width=4, fig.align='left'}

lower<-function(rc) {
 ut = upper.tri( rc$r , diag=T) 
 rc$r[ut] = 0
 rc$P[ut] = NA
 return(rc)
}

sig.rc<-function(rc) {
  sig = rc$P > 0.05
  rc$r[sig] = 0
  return(rc)
}

cors = list()
for (name in names(patient.data)) {

  sample.normalised.segvals = patient.data[[name]]$norm.seg.vals # prog
  sample.normalised.segvals =  sample.normalised.segvals[, patient.data[[name]]$info$Samplename ]

  colnames(sample.normalised.segvals) =  paste(patient.data[[name]]$info$Endoscopy.Year,  c(1:nrow(patient.data[[name]]$info)) , sep='.')

  #sample.normalised.segvals = patient.data$AD0361$norm.seg.vals # non-p
  rc = rcorr(as.matrix(sample.normalised.segvals), type='pearson')
  #rc = sig.rc(lower(rc))
  rc = sig.rc(rc)

  diff = unlist(apply(rc$r, 1, function(x)  sum(x[x < 0])-1 ))/nrow(rc$r)
  sim = unlist(apply(rc$r, 1, function(x)  sum(x[x > 0])-1 ))/nrow(rc$r)

  t.test(abs(sim), abs(diff))

  cors[[name]][['sim']] = sim
  cors[[name]][['diff']] = diff
}

unlist(lapply(cors, function(x) t.test(abs(x$sim), abs(x$diff))$p.value))

simPNP = wilcox.test(unlist(lapply(cors[subset(sum.patient.data, Status == 'P')$Patient], function(x) sum(x$sim))),
            unlist(lapply(cors[subset(sum.patient.data, Status == 'NP')$Patient], function(x) sum(x$sim))))

diffPNP = wilcox.test(unlist(lapply(cors[subset(sum.patient.data, Status == 'P')$Patient], function(x) sum(x$diff))),
            unlist(lapply(cors[subset(sum.patient.data, Status == 'NP')$Patient], function(x) sum(x$diff))))
```

There's no clear difference (currently) between P and NP for the similar correlation (`r simPNP$p.value`) or the inverse (`r diffPNP$p.value`) correlations.  There are other ways we may be able to look at this still.

## Look at Cancer Consensus Genes in the variable regions

```{r, hmm, echo=F, message=F, warning=F}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)

noLoc = grep(':-',ccgenes$Genome.Location)
noLocGenes = ccgenes[noLoc,]

ccgenes = ccgenes[-noLoc,]

locs = as.data.frame(do.call(rbind, strsplit(ccgenes$Genome.Location, ':|-')))
colnames(locs) = c('chr','start','end')
locs[c('start','end')] = lapply(locs[c('start','end')], function(x) as.numeric(as.character(x)))

ccgenes = cbind(locs, ccgenes)
ccgenes = makeGRangesFromDataFrame(ccgenes, keep.extra.columns = T, start.field = 'start', end.field = 'end')

# ov = findOverlaps(ccgenes, ccgenes)
# ov = ov[queryHits(ov) != subjectHits(ov)]
# ccgenes = ccgenes[-subjectHits(ov)[1]]

cancer.consensus.genes<-function(reg, ccgenes) {
  reg = as_tibble(reg)
  gr = GenomicRanges::makeGRangesFromDataFrame(reg, start.field = 'start.pos', end.field = 'end.pos')

  ov = findOverlaps(ccgenes, gr)
  reg$Gene.Symbols = NA
  for (hit in unique(subjectHits(ov))) {
    reg[hit, 'Gene.Symbols'] =  paste(ccgenes[queryHits(ov)[which(subjectHits(ov) == hit)],]$Gene.Symbol, collapse=',')
  }

  rwg = cbind(reg, reg[,-(1:5)])
  return(rwg)
}

rm(var.regions)
for (patient.name in names(patient.data)) {
  print(patient.name)
  regions = patient.data[[patient.name]]$variable.regions[,c('chrom','arm','start.pos','end.pos')]
  if (nrow(regions) <= 0) next

  reg = cancer.consensus.genes(regions, ccgenes)
  x = patient.data[[patient.name]]$variable.regions[which(!is.na(reg$Gene.Symbols)), patient.data[[patient.name]]$info$Samplename]

  gl = apply(x, 1, function(y)  ifelse(length(which(y >= 1.1)) > length(which(y <= 0.9)), 'gain', 'loss') )
  reg[which(!is.na(reg$Gene.Symbols)), 'CN'] = gl
  reg = reg[,c('chrom','arm','start.pos','end.pos','CN','Gene.Symbols')]

  for (i in 1:nrow(reg)) {
    if (is.na(reg[i,'Gene.Symbols'])) next
    genes = unlist(strsplit(reg[i,'Gene.Symbols'], ','))

    reg[i,'Gene.Symbols'] = genes[1]

    if (length(genes) > 1)
      reg = rbind(reg, cbind(reg[i, (1:5)], 'Gene.Symbols' = genes[2:length(genes)]))
  }
  reg = arrange(reg, chrom, arm, start.pos)

  status = subset(sum.patient.data, Patient == patient.name)$Status

  reg = cbind(patient.name, status, reg)

  if (!exists('var.regions')) {
    var.regions = reg
  } else {
    var.regions = rbind(var.regions, reg)
  }
}

var.regions = var.regions[!is.na(var.regions$Gene.Symbols),]
var.regions$patient.name = as.character(var.regions$patient.name)

var.regions = var.regions[-grep('CDKN2A\\(p14\\)', var.regions$Gene.Symbols),]

gene.cnt = var.regions %>%
  group_by(Gene.Symbols, CN) %>%
  summarise(Progressors=length(which(status == 'P')), Non.Progressors=length(which(status == 'NP')) )
gene.cnt[,unique(var.regions$patient.name)] = 0
gene.cnt = gene.cnt[which(!is.na(gene.cnt$Gene.Symbols)),]

for (i in 1:nrow(gene.cnt)) {
  gene = as.character(gene.cnt[i,'Gene.Symbols'])
  type = as.character(gene.cnt[i,'CN'])
#  print(paste(gene, type))
#  print(subset(var.regions, Gene.Symbols == gene & CN == type)$patient.name)
  gene.cnt[i,unique(subset(var.regions, Gene.Symbols == gene & CN == type)$patient.name)] = 1
}

gene.cnt = arrange(gene.cnt, -Progressors)
gene.p = subset(gene.cnt, Progressors >= Non.Progressors)

m = melt(gene.p[1:50,c(1:4)], id.vars=c('Gene.Symbols', 'CN'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
m$CN = factor(m$CN)

p = ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')
p + geom_point(data=m[m$CN == 'gain' & m$value > 0,], aes(label='gain',y=Gene.Symbols, x=variable), size=2, shape=24, fill='red', color='red') +
  geom_point(data=m[m$CN == 'loss' & m$value > 0,], aes(label='loss',y=Gene.Symbols, x=variable), color='green4', size=2, shape=25, fill='green4')

gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
m = melt(gc[1:50,], id.vars=c('Gene.Symbols'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')

@@

## Is HGD over represented for variable regions?
```{r, hgd, echo=F, message=F, warning=F}

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  print(name)
  sn = grep('HGD', patient.data[[name]]$info$Pathology)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next
  samplenames = patient.data[[name]]$info$Samplename

  ## HGD
  pvnp[name, 'hgd'] = length(which(patient.data[[name]]$variable.regions[samplenames[sn]] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[samplenames[sn]] <= loss.threshold))

  # BE-LGD
  pvnp[name, 'be'] = length(which(patient.data[[name]]$variable.regions[samplenames[-sn]] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[samplenames[-sn]] <= loss.threshold))
}

# No difference

tt = t.test(pvnp$hgd, pvnp$be)
pvnp = cbind(pvnp, (pvnp %>% rowwise() %>% mutate(total=hgd+be, hgd.ratio=round(hgd/(hgd+be),3) ))[c('total','hgd.ratio')])

pander(pvnp, justify='left', 
       caption=paste("Variable regions found in HGD vs non-HGD samples of progessor patients.  There's no significant difference between them (t-test p=",signif(tt$p.value,2),').  Not all patients have HGD samples (NAs).',sep=''))
